Efficient static analysis of Marlowe contracts

### 2020-05-20 papers/static_analysis2020/paper.tex
More rewording and fixes

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Account identifiers include an account owner}; an account owner is
a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{Classification of money and tokens into accounts} separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully, it will still wait for a null deposit to
be made, and it will continue as if everything is correct, this way
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested When constructs in the contract, since only one input
per When can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This prevents DoS attacks, and makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

The SBV library supports implementing Haskell functions in a way that
the same implementation can be used:
\begin{itemize}
\item With concrete parameters, as a normal Haskell function
\item With symbolic variables, so that properties can be check for satisfiability
using an SMT solver
\item As part of QuickCheck properties, for testing randomly.
\end{itemize}
Parameters that can be used symbolically are wrapped in a monad called
\texttt{SBV}. Functions that depend on symbolic parameters also return
values wrapped in the \texttt{SBV} monad.

@@

Unfortunately, there are a couple of issues with this approach that
we review in Section~\ref{subsec:Considerations}.

\subsection{Considerations\label{subsec:Considerations}}

SBV does not currently support complex custom data types. It has some
support for nonary data types, but not for data types with parameters.
It does support the \texttt{Either} type, and the \texttt{Tuple} type.
So it is possible to have conversions between custom data types and
nested \texttt{Either} and \texttt{Tuple} type synonyms. We originally
implemented this conversion using Template Haskell, since by using
this technique the symbolic implementation can remain similar to the
concrete one.

@@

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can calculate limits for:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data we can deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} are
called.
\item The number of accounts that the contract will use, and an
the number of elements in each of the associative maps
that comprise the \texttt{State} of the contract.
\end{itemize}
If we keep the \texttt{Contract} parameter of the semantics concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for \texttt{Close} we know we
just have to call it as many times as the maximum number of accounts
there can possibly be, since every time we call it it will refund
one of the accounts.

@@

However, there is one more problem. The output \texttt{Contract} returned
by the function, which is wrapped inside the \texttt{TransactionOutput},
is symbolic, since it depends on the current \texttt{TransactionInput}
and \texttt{State}, which are both symbolic.

We get around this problem by using a continuation style. Instead
of returning the \texttt{TransactionOutput} directly, we take a continuation
function that takes the concrete contract and the symbolic version
of \texttt{TransactionOutput} without the contract.

@@

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain
intuitive and close to the original semantics and model that we are
trying to analyze, as well as making changes less likely to introduce
errors than a more serious restructure.

\subsubsection{Removing unnecessary parts from the analysis }

This is probably the most straightforward optimization we can make.
When we use the same or similar code for both the analysis and the
implementation, we may end up including code that is not relevant
to the analysis.

@@

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment
(we have also proven this using Isabelle\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy}
(last visited on 04/04/2020)}): it only pays whatever there is available.
So we can safely remove it from the analysis.

\subsubsection{Avoiding high level abstractions }

@@

For example, in the case of Marlowe static analysis, we initially
implemented a symbolic associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this is very costly in a
solver: searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

@@

\subsubsection{Reducing search space by normalising parameters\label{subsec:Reducing-search-space-by-normalising-parameters}}

@@

However, we can devise a normal form for input traces, in which there
is a maximum of one input per transaction. We only need to make sure
that, for every trace, if it produces a warning, there exists a trace
with only one input per transaction that also produces a warning.
Using the automated proof assistant Isabelle \cite{nipkow2002isabelle},
we have shown the stronger property that splitting a transaction in
two does not modify the effect on a contract as long as the transaction
does not produce an error\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/SingleInputTransactions.thy}
(last visited on 29/04/2020)}.

This optimization reduces the search space considerably, but transactions
may still have no inputs, so it still leaves room for many possible
ways of combining the inputs in transactions. We explain how we reduced
the search space even further in the next section.

\subsection{Heavyweight modifications \label{subsec:Heavyweight-modifications}}

@@

\subsubsection{Reducing search space by using normalised execution paths relevant
to the property \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}

@@

Unfortunately, having static analysis be very different from implementation
makes the approach much more error prone, and in making an argument
for the trustworthiness of the analysis there are many more assumptions
that need to be made and reasoned about. In Section~\ref{sec:Testing}
we explore ways of mitigating this problem.

@@

One subtle detail that we realised during testing is that, even though
the diagram above shows that a \texttt{When} belongs to the new transaction,
because it will still be there after the previous transaction is executed,
the observations in the \texttt{Notify} cases will correspond to the
information there was in the \texttt{State} before the \texttt{When}
was executed, with the exception of the slot interval, which will still be
considered to be that of the transaction that triggers the \texttt{When} (the
new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
cases, since the amount to deposit must be calculated without considering the
effects of the deposit itself, we need to refer to the \texttt{State} before
the deposit was made.

\subsubsection{Minimizing the representation of inputs and outputs }

When we initially implemented the efficient version of static analysis
for Marlowe, we did not pay any attention to the inputs and outputs.
The first version we wrote of the SBV property would not take any
symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
as input, and it would return a symbolic boolean that said whether
the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
turns out to be invalid, we also want to know why, so we later modified
the property to give a counterexample that illustrated what went wrong.
The original implementation still used some intermediate symbolic
variables, but they were anonymous, and they were created during the
exploration of the contract.

@@

In order to translate this sequence of numbers into a proper list
of transactions that is human readable and we can use to report the
counter example, we need to iterate through the list and see the evolution
of the contract with each transaction, using the concrete semantics.
This way, we can gather the rest of necessary information, such as
whether the transaction input is a \texttt{Deposit}. We can also see
which of the transactions does not have an effect, i.e: they produce
\texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

@@

In order to avoid these problems, we use a mixture of automated proof
assistants to verify our assumptions, and property based testing to
test for consistency and equivalence.

@@

In all the contracts, the implementation that has the heavyweight optimisations
performs much better and scales further. In the case of the \textit{auction}
(Table~\ref{table1}) and \textit{crowdfunding} (Table~\ref{table2}) contracts,
whose size grows exponentially, both approaches quickly overwhelm the resources
available in the execution environment. On the other, in the case of \textit{rent}
(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}), which grow
linearly, by using the lightweight version the problem becomes intractable much
faster than with the heavyweight version.

@@

These results suggest that SBV library is able to generate constraints in a way
that they are handled efficiently by Z3, but the process itself can be costly
if we are not careful. Nevertheless, the overall execution time required by Z3
is also lower in the case of the version with heavyweight optimisations and it
grows more slowly, which suggests the optimisations described in this paper
affect both stages of the process.

@@

Another advantage of the optimised implementation is that, because
it relies on fewer and simpler features, it is compatible with more
SMT solvers which, in turn, it means that it is less reliant on the
correctness of a single SMT solver. If an SMT solver fails to give
an answer, we can try a different one; if we want to ensure that a
contract is valid, we can test it with several SMT solvers and ensure
all of them agree.

>>>>

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Classification of money and tokens into accounts} separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{Account identifiers include an account owner}. An account owner
is a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully: it will still wait for a null deposit to
be made, and it will continue as if everything is correct; this way,
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested \texttt{When} constructs in the contract, since only one input
per \texttt{When} can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This bound prevents DoS attacks, and it makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

The SBV library supports implementing Haskell functions in a way that
the same implementation can be used:
\begin{itemize}
\item With concrete parameters, as a normal Haskell function
\item With symbolic variables, so that properties can be checked for satisfiability
using an SMT solver
\item As part of QuickCheck properties, for random testing.
\end{itemize}
Parameters that can be used symbolically are wrapped in a monad called
\texttt{SBV}. Functions that depend on symbolic parameters also return
values wrapped in the \texttt{SBV} monad.

@@

Unfortunately, there are a couple of issues with this approach that
we review in Section~\ref{subsec:considerations}.

\subsection{Additional considerations\label{subsec:considerations}}

SBV does not currently support complex custom data types. At the time
of writing, SBV has some support for nonary data types, but not for
data types with parameters.
It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
Thus, it is possible to have an isomorphism between custom data types and
nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
implementation automatically generates conversion functions between these
by using Template Haskell. By using this technique the symbolic implementation
can remain similar to the concrete one.

@@

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can easily infer:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data, we can also deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
be called.
\item The number of accounts that the contract will use, and an
the number of elements there may be in each of the associative maps
that comprise the \texttt{State} of the contract at any given point
of the execution.
\end{itemize}
If we keep the \texttt{Contract} parameter of the semantics concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for \texttt{Close} we know we
just have to call it as many times as the maximum number of accounts
there can possibly be, since every time we call it it will refund
one of the accounts.

@@

However, there is one more problem: the output \texttt{Contract} returned
by the function, which is wrapped inside the \texttt{TransactionOutput},
is symbolic, since it depends on the current \texttt{TransactionInput}
and \texttt{State}, which are both symbolic.

We get around this problem by using a continuation style. Instead
of returning the \texttt{TransactionOutput} directly, we take a continuation
function that takes the concrete \texttt{Contract} and the symbolic version
of \texttt{TransactionOutput} without the \texttt{Contract}.

@@

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain
intuitive and close to the original semantics and model that we are
trying to analyze, as well as making changes less likely to introduce
errors than a more serious restructuring.

\subsubsection{Removing unnecessary parts from the analysis.}

This is probably the most straightforward optimization to implement.
When we use the same or similar code for both the analysis and the
implementation, we may end up including code that is not relevant
to the analysis.

@@

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment
(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} using Isabelle \cite{nipkow2002isabelle})
(last visited on 04/04/2020)}: it only pays whatever there is available.
So we can safely remove it from the analysis.

\subsubsection{Avoiding high level abstractions.}

@@

For example, in the case of Marlowe's static analysis, we initially
implemented a symbolic associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this can be very costly:
searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

@@

\subsubsection{Reducing search space by normalising parameters.
\label{subsec:Reducing-search-space-by-normalising-parameters}}

@@

However, we can devise a normal form for input traces, in which there
is a maximum of one input per transaction. We only need to make sure
that, for every trace, if it produces a warning, there exists a trace
with only one input per transaction that also produces a warning.
Using the automated proof assistant Isabelle \cite{nipkow2002isabelle},
we have shown that, indeed, splitting transactions into transactions
with single inputs and the same slot interval as the original transactions
does not modify the effect of those transactions on a contract\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/SingleInputTransactions.thy}
(last visited on 29/04/2020)}.

This optimization reduces the search space considerably, but transactions
may still have no inputs, so it still leaves room for many possible
ways of combining the inputs in transactions. We explain how we reduced
the search space even further in Section~\ref{subsec:Heavyweight-modifications}.

\subsection{Heavyweight modifications. \label{subsec:Heavyweight-modifications}}

@@

\subsubsection{Reducing search space by using normalised execution paths relevant
to the property. \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}

@@

Unfortunately, having static analysis be very different from implementation
makes the approach much more error prone, and in making an argument
for the trustworthiness of the analysis there are many more assumptions
that need to be made and reasoned about. In Section~\ref{sec:Testing},
we explore ways of mitigating this problem.

@@

One subtle detail that we realised during testing is that, even though
the diagram above shows that a \texttt{When} belongs to the new transaction,
because the \texttt{When} will not be consumed by the previous transaction,
\texttt{Observation}s in the \texttt{Notify} cases will correspond to the
information there was in the \texttt{State} before the \texttt{When}
was executed, with the exception of the slot interval, which will still be
considered to be that of the transaction that triggers the \texttt{When} (the
new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
cases, since the amount to deposit must be calculated without considering the
effects of the deposit itself, we need to refer to the \texttt{State} before
the deposit was made.

\subsubsection{Minimizing the representation of inputs and outputs.}

When we initially implemented the efficient version of static analysis
for Marlowe, we did not pay any attention to the inputs and outputs.
The first version we wrote of the SBV property would not take any
symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
as input, and it would return a symbolic Boolean that determined whether
the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
turns out to be invalid, we will also want to know why, so we later modified
the property to give a counterexample that illustrated what went wrong.
The original implementation still used some intermediate symbolic
variables, but they were anonymous, and they were created during the
exploration of the contract.

@@

In order to translate this sequence of numbers into a proper list
of transactions that is human readable and we can use to report the
counter example, we need to iterate through the list and see the evolution
of the contract with each transaction, using the concrete semantics.
This way, we can gather the rest of necessary information, such as
whether the transaction input is a \texttt{Deposit}. We can also see
which of the transactions do not have an effect, i.e: they produce
\texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

@@

In order to avoid these problems, we combine the use of automated proof
assistants to verify our assumptions and the use of property based testing to
test for consistency and equivalence.

@@

In all the contracts, the implementation that has the heavyweight optimisations
performs much better and scales further. In the case of the \textit{auction}
(Table~\ref{table1}) and \textit{crowdfunding} (Table~\ref{table2}) contracts,
whose size grows exponentially, both approaches quickly overwhelm the resources
available in the execution environment. On the other hand, in the case of \textit{rent}
(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}) contracts, which grow
linearly, by using the lightweight version, the problem becomes intractable much
faster than with the heavyweight version.

@@

These results suggest that SBV library is able to generate constraints in a way
that they are handled efficiently by Z3, but the process itself can be costly
if we are not careful. Nevertheless, the overall execution time required by Z3
is also lower in the case of the version with heavyweight optimisations and it
grows more slowly, which suggests the optimisations described in this paper
affect both stages of the process to some extent.

@@

Another advantage of the optimised implementation is that, because
it relies on fewer and simpler features, it is compatible with more
SMT solvers which, in turn, means that it is less reliant on the
correctness of a single SMT solver. If an SMT solver fails to give
an answer, we can try a different one; if we want to ensure that a
contract is valid, we can test it with several SMT solvers and ensure
all of them agree.
